Project assignment - Tahoe-LAFS with SHA-3 candidates

### 2010-12-11 paper/main.tex
Updated disposition on Discussion chapter

@@

@@

The chosen implementations are shown in Table \ref{tbl:sha3:implementations}.
Their compiler flags used, which are extracted from the \ac{SUPERCOP}
results, can be seen in Table \ref{tbl:sha3:compilerflags}.

It should be noted that \emph{32} in blake32 does not refer to the output length
of the function, but the fact that the algorithm is optimized for 32-bit usage.
Similarly, \emph{512} in keccakc512 does not mean 512-bit output, but 512 cycles in the
internal function.

@@

\paragraph{JH} seems to be broken in every implementation with regard to
Criterion 2. For both the optimized versions found in \ac{SUPERCOP} and the
reference implementations in the \ac{NIST} submission package, Hash("foofoo")
will not yield the same result as two consecutive calls to Update("foo").
However the implementation in sphlib does fulfill this requirement. We have
contacted the author, Thomas Pornin, which confirms this problem with the
official JH candidate and that the author has been made aware of the problem
but has not released any updated version yet (personal communication, October
16, 2010).

@@

Since Tahoe-\ac{LAFS} is written in Python, we needed some way of interacting
with the C implementations of the NIST candidates. We did this by the use of
Cython, and built a python object-oriented library which we named SHA3lib.
SHA3lib supports all of the 14 second round \ac{SHA}-3 candidates, but only in
their 256-bit form.

\subsection{SHA3lib}
SHA3lib is built to mimic the functionality of
hashlib\footnote{\url{http://docs.python.org/library/hashlib.html}}, the
default python library for MD5, \ac{SHA}-1 and the \ac{SHA}-2 family. This
means that we built class names for each candidate on the form
candidatename+256 and that the classes included the following functions:

\begin{itemize}
    \item Constructur with an optional data element that does the same as
    Update()
    \item Update() takes data that should be considered in hashing
    \item digest() that produce a raw byte-digest from submitted data
    \item hexdigest() that produce a hex representation of the digest
    \item copy() that copies the internal state of the function and produces a
    copy of the object.
\end{itemize}

@@

Basically the class echo resides in Cython and in C and is responsible for
bringing the C functionality of the C implementation to Python. By compiling
echo it will be made available for use in Python.  The class echo256 is a
simple python class which inherits from echo. The only action this class
actually takes is to set the output size to 256-bit. If the underlying
C-functions would have supported multiple output sizes the same could have been
done to incorporate those.

\subsubsection{Cython}
We employ Cython to generate bindings for the C implementation and the
different candidates, as well as to create the base python classes. To access
functions in C we need to generate a Cython header file, which pretty much does
the same as a C header file; it tells which functions and data types are
available. How we accomplish this for an implementation that satisfies the
\ac{NIST} \ac{API} is shown in Listing \ref{lst:cython_header}.

@@

We then create a wrapper-class that mimics hashlib, this is shown in Listing
\ref{lst:cython_impl}. Functions declared with def er python functions,
{\verb cdef } are Cython only while {\verb cpdef } are functions that can be
accessed from both. While this is just the implementation for the echo
candidate, the almost exact same pattern is used for all of the other
candidates.

@@

\subsubsection{\ac{Distutils}}
We use the \ac{Distutils}\footnote{\url{http://docs.python.org/distutils/}} to
build and install our code. The alternative to doing this would be to manually
transform Cython files to C-files, build the shared library from this and other
relevant C-files and then install by copying into the correct system directory
where python libraries should be stored. \ac{Distutils} also support expansion
through self defined commands, something we have utilised to do Unit Testing.
Options and configuration for \ac{Distutils} is done through a file
{\verb setup.py}. Relevant configuration for compiling the echo extension, and
installing SHA3lib can bee seen in Listing \ref{lst:setup_py}.

@@

We see in Tables \ref{tbl:supercop:long}, \ref{tbl:supercop:4096},
\ref{tbl:supercop:1536}, \ref{tbl:supercop:576}, \ref{tbl:supercop:64} and
\ref{tbl:supercop:8} that in general BLAKE, \ac{BMW}, Luffa, Shabal and Simd
outperform \ac{SHA}-256. However, given that Tahoe-LAFS uses double
\ac{SHA}-256, this means that it will also have to hash another 32 bytes for
every message it returns the digest for. Since \ac{SUPERCOP} does not have
benchmark for 32 byte messages, we will have to estimate the amount of cycles
required for this.  If we assume that cycles/bytes is linear from 8 byte to 64
byte, we get the following estimate for the speed of hashing 32 bytes:

$$179.62 - \frac{179.62 - 38.12}{64 - 8} * (32-8)= 118.98 (Cycles/byte)$$

We can then estimate the result of each message by the following formula:
$$\frac{median * size + 118.98*32}{size} = SHA256d$$ \emph{estimate}

@@

Estimates based on this calculation are shown in Table
\ref{tbl:supercop:sha256d}. What we can see that for lower lengths of data,
more of the candidates are faster than double \ac{SHA}-256. However, given the
practical situation that Tahoe-\ac{LAFS} is a file system, most files processed
will be larger than the file sizes estimated here. Even though not all
hashing operations done in Tahoe are related to the file size, we expect that
only BLAKE, \ac{BMW}, Luffa, Shabal and SIMD will be able to outperform double
\ac{SHA}-256.

@@

\section{Exclusion of mutable file testing}
% Implications of only testing immutable files? (mutable files also
%   include directories)
% Adding a capability to support new hash functions in tahoe. (Our
%   current way breaks backwards compability)

@@

To be able to extend the Tahoe-LAFS with new functionality to fully support new
hash functions, one have to add a new capability (Zooko O'Whielacronx, personal
communication, September 29, 2010). This ensures backwards compatibility and can
be an optional feature for users to employ. However, we found that adding a new
capability was out of scope for this assignment.

% This chapter should contain (in listed and grouped order):
% -----------------------------------------------------------
% Describe inconsistencies in the numbers, and try to explain them.
% Time spent not hashing varies because: Variable use of CPU regs?
%
% Compare our results with general (SUPERCOP) results.
% Optimized candidates - which gain could be expected from more optimized
%   candidates?
% How much gain can Tahoe-LAFS get by switching to SHA-3?
%
% What if we had used another coding scheme? - does Cython lower the
%   candidates performance?
%
% More general stuff. What could have been done differently? Criticize
% ourselves.

>>>>

%  Writers should follow a checklist consisting of:
% Motivation: Why do we care about the problem and results?
% Problem Statement: What problem are we trying to solve? Scope/limits.
% Approach: How did we go about solving or making progress on the problem?
% Results: What is the answer? Numbers, not vague 'very', 'small' etc.
% Conclusions: What are the implications of your answer? Further work.
%
%  Each section is typically a single sentence, although there is room for
%  creativity.

@@

The work behind this project report was carried out during the autumn semester
in 2010 at the Norwegian University of Science and Technology (NTNU), Department
of Telematics (ITEM).
\vspace{13pt}

We would like to thank our fellow classmates for interesting discussions and a
lot of fun as the time and semesters fly by.
\vspace{13pt}

Secondly, we would like to state our appreciation to our supervisor, Danilo
Gligoroski, and his coworker on Blue Midnight Wish, Rune E. Jensen, for
providing useful feedback and interesting stories from the cryptographic
community.
\vspace{13pt}

Lastly, the Open Source community behind Tahoe-LAFS deserves credit for the
immense amount of hours they put on an awesome project, and Thomas Pornin for
providing a working copy of his library of implementations, which saved us a lot
of work.

@@

The chosen implementations are shown in Table \ref{tbl:sha3:implementations}.
Their compiler flags used, which are extracted from the \ac{SUPERCOP} results,
can be seen in Table \ref{tbl:sha3:compilerflags}. Compiler flags are described
in the GCC manual\footnote{GCC Manual \url{http://gcc.gnu.org/onlinedocs/}}.
Compiler flags used for every implementation which might affect performance are:

\begin{itemize}
    \item -fomit-frame-pointer
    \item -mssse3
    \item -pthread
    \item -fPIC
\end{itemize}

It should be noted that \emph{32} in blake32 means the function works on 32-bit
words, and output a 32-byte digest (256 bit). Similarly, \emph{512} in keccakc512 
does not mean 512-bit output, but 512 bits in the internal capacity.

@@

\paragraph{JH} seems to be broken in every implementation with regard to
Criterion 2. For both the optimized versions found in \ac{SUPERCOP} and the
reference implementations in the \ac{NIST} submission package,
{\verb Hash("foofoo") } will not yield the same result as two consecutive calls
to {\verb Update("foo") }.  However the implementation in sphlib does fulfill
this requirement. We have contacted the author, Thomas Pornin, which confirms
this problem with the official JH candidate and that the author has been made
aware of the problem but has not released any updated version yet
\cite{s_pornin}.

@@

Since Tahoe-\ac{LAFS} is written in Python, we needed some way of interacting
with the C implementations of the NIST candidates. We did this by the use of
Cython, and built a Python object-oriented library which we named SHA3lib.
SHA3lib supports all of the 14 second round \ac{SHA}-3 candidates, but only in
their 256-bit form.

\subsection{SHA3lib}
SHA3lib is built to mimic the functionality of
hashlib\footnote{\url{http://docs.python.org/library/hashlib.html}}, the
default Python library for MD5, \ac{SHA}-1 and the \ac{SHA}-2 family. This
made us create class names for each candidate on the form
\emph{candidate name+256}. In addition, the classes included the following
functions:

\begin{itemize}
    \item Constructor with an optional data element that does the same as
    {\verb update() }
    \item {\verb update() } takes data that should be considered in hashing
    \item {\verb digest() } that produce a raw byte-digest from submitted data
    \item {\verb hexdigest() } that produce a hex representation of the digest
    \item {\verb copy() } that copies the internal state of the function and
        produces a copy of the object.
\end{itemize}

@@

Basically the class \textbf{echo} resides in Cython and in C, and is responsible
for bringing the C functionality of the C implementation to Python. By compiling
\textbf{echo} it will be made available for use in Python.  The class
\textbf{echo256} is a simple Python class which inherits from \textbf{echo}. The
only action this class actually takes is to set the output size to 256-bit. If
the underlying C-functions would have supported multiple output sizes the same
could have been done to incorporate those.

\subsubsection{Cython}
We employ Cython to generate bindings for the C implementation and the
different candidates, as well as to create the base Python classes. To access
functions in C we need to generate a Cython header file, which pretty much does
the same as a C header file; it tells which functions and data types are
available. How we accomplish this for an implementation that satisfies the
\ac{NIST} \ac{API} is shown in Listing \ref{lst:cython_header}.

@@

We then create a wrapper-class that mimics hashlib, this is shown in Listing
\ref{lst:cython_impl}. Functions declared with def er Python functions,
{\verb cdef } are Cython only while {\verb cpdef } are functions that can be
accessed from both. While this is just the implementation for the echo
candidate, the almost exact same pattern is used for all of the other
candidates.

@@

\subsubsection{\ac{Distutils}}
We use \ac{Distutils}\footnote{\url{http://docs.python.org/distutils/}} to
build and install our code. The alternative to doing this would be to manually
transform Cython files to C-files, build the shared library from this and other
relevant C-files and then install by copying into the correct system directory
where Python libraries should be stored. \ac{Distutils} also support expansion
through self defined commands, something we have utilised to do Unit Testing.
Options and configuration for \ac{Distutils} is done through a file
{\verb setup.py}. Relevant configuration for compiling the echo extension, and
installing SHA3lib can bee seen in Listing \ref{lst:setup_py}.

@@

We see in Tables \ref{tbl:supercop:long}, \ref{tbl:supercop:4096},
\ref{tbl:supercop:1536}, \ref{tbl:supercop:576}, \ref{tbl:supercop:64} and
\ref{tbl:supercop:8} that in general BLAKE, \ac{BMW}, Luffa, Shabal and Simd
outperform \ac{SHA}-256. However, given that Tahoe-LAFS uses double
\ac{SHA}-256, this means that it will also have to hash another 32 bytes for
every message it hashes. Since \ac{SUPERCOP} does not have benchmark for 32
byte messages, we will have to estimate the amount of cycles required for this.
If we assume a worst case scenario of hashing 32 bytes, where the cycles/byte
count is exactly the same as for 8 bit median we get:

$$\frac{median * size + 179.62*32}{size} = SHA256d estimate$$

@@

Estimates based on this calculation are shown in Table
\ref{tbl:supercop:sha256d}. What we can see that for lower lengths of data,
more of the candidates are faster than double \ac{SHA}-256. However, given the
practical situation that Tahoe-\ac{LAFS} is a file system, most files processed
will be larger than the file sizes estimated here. Not all hashing
operations in Tahoe-\ac{LAFS} are strictly dependant on the original file size,
which would mean that all candidates would have a chance of outperforming
double \ac{SHA}-256. However this estimate is worst case, which leads us to
believe that at least for larger input sizes (over 1536) only the five fastest
candidates will have a change of outperforming double \ac{SHA}-256.

@@

%TODO: Vi har andre compiler argumenter - dette må diskuteres

\section{Exclusion of mutable file testing}
% Implications of only testing immutable files? (mutable files also
%   include directories)
% Adding a capability to support new hash functions in tahoe. (Our
%   current way breaks backwards compatibility)

@@

To be able to extend the Tahoe-LAFS with new functionality to fully support new
hash functions, one have to add a new capability \cite{t_zooko}. This ensures
backwards compatibility and can be an optional feature for users to employ.
However, we found that adding a new capability was out of scope for this
assignment.

\section{Thoughts on the measurements}

% Describe inconsistencies in the numbers, and try to explain them.
% Time spent not hashing varies because: Variable use of CPU regs?
%
% Compare our results with general (SUPERCOP) results.
%
% How much gain can Tahoe-LAFS get by switching to SHA-3?
%
% Drag in the five selected remaining candidates.

\section{Improvement of methods}
% More general stuff. What could have been done differently? Criticize
% ourselves.
%
% Optimized candidates - which gain could be expected from more optimized
%   candidates?

\subsection{The use of Cython}
% What if we had used another coding scheme? - does Cython lower the
%   candidates performance?

The Python-bindings generated for Cython in C are not necessarily the most
optimized versions possible to write and they are ugly to read or change (from
a C perspective). However we consider them to be good enough, and they are
equal for all the \ac{SHA}-3 candidates which means they should give the same
performance penalty. The \ac{SHA}-256 version will of course not suffer from
the same penalty, however we believe the possible performance gain is small.
This is due to the fact that no intensive operations take place in our
Cython-code.

The alternative would be creating manual Python-bindings for all the candidates
which would require a lot more work on our side, especially given the lack of a
good enough and common \ac{API} for the implementations in \ac{SUPERCOP}.
Another fact important for this decision was that we did not know exactly which
implementation of each candidate we would end up using, and we ended up trying
several for most candidates.
